Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-08-28 Documents/M1/realizableFermionicM1/sections/realizability.tex
reworked section due to comments from Cory

Our goal is to simulate massless fermions (e.g., neutrinos) and study their interactions with matter.  
The principal objective is to obtain the fermionic distribution function $f$ (or moments of $f$ as in the two-moment model employed here).  
The Pauli exclusion principle requires the distribution function to satisfy the condition $0 \le f \le 1$, which puts restrictions on the admissible values for the moments of $f$.  
In this paper, we seek to design a numerical method for solving the system of moment equations given by Eq.~\eqref{eq:momentEquations} that preserves realizability of the moments; i.e., the moments evolve within the set of admissible values as dictated by Pauli's exclusion principle.  
(Since we are only concerned with the angular dependence of $f$ in this section, we simplify the notation by suppressing the $\vect{z}$ and $t$ dependence and write $f(\omegaNu,\vect{z},t)=f(\omegaNu)$.)

We begin with the following definition of moment realizability.  
\begin{define}
  The moments $\vect{\cM}=\big(\cJ,\vect{\cH}\big)^{T}$ are realizable if they can be obtained from a distribution function satisfying $0 < f(\omegaNu) < 1~\forall~\omegaNu\in\bbS^{2}$.
  \label{def:momentRealizability}
\end{define}
\begin{rem}
  As in \cite{lareckiBanach_2011}, in Definition~\ref{def:momentRealizability}, and in the rest of this paper, we exclude the cases $f=0$ and $f=1$ almost everywhere (a.e.) on $\bbS^{2}$, which would give $\cJ=0$ and $\cJ=1$, respectively.  
\end{rem}

Realizable moments satisfy algebraic constraints.  
We proceed by stating the constraints on the moments by restating results from Theorem~7.1 in \cite{banachLarecki_2017a} in the following Lemma.  
The proof is given in \cite{banachLarecki_2017a}.  
(See also \cite{lareckiBanach_2011,banachLarecki_2013}.)
\begin{lemma}
  Suppose that $0 < f(\omegaNu) < 1~\forall~\omegaNu\in\bbS^{2}$.  
  Then the moments $\vect{\cM}=(\cJ,\vect{\cH})^{T}$, defined as in Eq.~\eqref{eq:angularMoments}, satisfy $0 < \cJ < 1$ and $\big(1-\cJ\big)\,\cJ-|\vect{\cH}| > 0$. 
  \label{lem:MomentRealizable} 
\end{lemma}
%\begin{proof}
%  First, let $\mu=\cos\thetaNu$ and $\bbS^{1}:=\{\,\mu\,|\,\mu\in[-1,1]\,\}$, and write
%  \begin{equation}
%    \cJ=\f{1}{2}\int_{\bbS^{1}}\mathfrak{f}(\mu)\,d\mu,
%    \quad\text{where}\quad
%    \mathfrak{f}(\mu)=\f{1}{2\pi}\int_{0}^{2\pi}f(\mu,\phiNu)\,d\phiNu.  
%  \end{equation}  
%  Since $f(\omegaNu)\in[0,1]$, we have $\mathfrak{f}(\mu)\in[0,1]$, which implies $\cJ\in[0,1]$.  
%  Next we need to show that $|\vect{\cH}|\le(1-\cJ)\,\cJ$.  
%  Let $\widehat{\vect{n}}\in\bbR^{3}$ (independent of $\omegaNu$) be an arbitrary unit vector, and write
%  \begin{equation}
%    \widehat{\vect{n}}\cdot\vect{\cH}
%    =\f{1}{2} \int_{\bbS^{1}}\mathfrak{f}(\mu)\,\mu\,d\mu,
%    \label{eq:NdotH}
%  \end{equation}
%  where we have aligned the polar axis of the spherical momentum space coordinate system with $\widehat{\vect{n}}$; i.e., $\widehat{\vect{n}}\cdot\vect{\ell}(\omega)=\mu$.  
%  First, following \cite{banachLarecki_2017a}, we can write the integral on the right-hand side of Eq.~\eqref{eq:NdotH} as
%  \begin{equation*}
%    (1-\cJ)\cJ - \f{1}{2}\int_{-1}^{1-2\cJ}(1-2\cJ-\mu)\,\mathfrak{f}(\mu)\,d\mu
%    -\f{1}{2}\int_{1-2\cJ}^{1}(2\cJ-1+\mu)\,[1-\mathfrak{f}(\mu)]\,d\mu.  
%  \end{equation*}
%  Since $\mathfrak{f}\in[0,1]$, $(1-2\cJ-\mu)$ is nonnegative on the interval $\mu\in[-1,1-2\cJ]$, and $(2\cJ-1+\mu)$ is nonnegative on the interval $\mu\in[1-2\cJ,1]$, we have $\widehat{\vect{n}}\cdot\vect{\cH}\le(1-\cJ)\,\cJ$.  
%  In a similar manner, we can also write the integral on the right-hand side of Eq.~\eqref{eq:NdotH} as
%  \begin{equation*}
%    -(1-\cJ)\cJ-\f{1}{2}\int_{-1}^{2\cJ-1}(1-2\cJ+\mu)\,[1-\mathfrak{f}(\mu)]\,d\mu
%    -\f{1}{2}\int_{2\cJ-1}^{1}(2\cJ-1-\mu)\,\mathfrak{f}(\mu)\,d\mu.  
%  \end{equation*}
%  Then, since $\mathfrak{f}\in[0,1]$, $(1-2\cJ+\mu)$ is nonpositive on the interval $\mu\in[-1,2\cJ-1]$, and $(2\cJ-1-\mu)$ is nonpositive on the interval $\mu\in[2\cJ-1,1]$, we have $\widehat{\vect{n}}\cdot\vect{\cH}\ge-(1-\cJ)\,\cJ$.  
%  We therefore have $-(1-\cJ)\,\cJ\le\widehat{\vect{n}}\cdot\vect{\cH}\le(1-\cJ)\,\cJ$.  
%  Since this holds for any unit vector $\widehat{\vect{n}}\in\bbR^{3}$, we must have $|\vect{\cH}|\le(1-\cJ)\,\cJ$.  
%\end{proof}

@@

For the realizability-preserving scheme developed in Section~\ref{sec:realizableDGIMEX}, we state some additional results.  
Lemma~\ref{lem:explicitStep} is used to help prove the realizability-preserving property of explicit steps in the IMEX scheme, while Lemmas~\ref{lem:implicitStep} and \ref{lem:correctionStep} are used to prove realizability-preserving properties of implicit steps.  
\begin{lemma}
  Let $\big\{\cJ_{a},\vect{\cH}_{a},\vect{\cK}_{a}\big\}$ and $\big\{\cJ_{b},\vect{\cH}_{b},\vect{\cK}_{b}\big\}$ be moments defined as in Eq.~\eqref{eq:angularMoments} with distribution functions $f_{a}$ and $f_{b}$, respectively, such that $f_{a}(\omega),f_{b}(\omega)\in(0,1)\,\forall\,\omega\in\bbS^{2}$.  
  Let $\Phi^{\pm}(\vect{\cM})=\f{1}{2}\big(\vect{\cM}\pm\widehat{\vect{e}}\cdot\vect{\cF}(\vect{\cM})\big)$, where $\widehat{\vect{e}}\in\bbR^{3}$ is an arbitrary unit vector, and $\widehat{\vect{e}}\cdot\vect{\cF}(\vect{\cM})=\big(\widehat{\vect{e}}\cdot\vect{\cH},\widehat{\vect{e}}\cdot\vect{\cK}\big)^{T}$.  
  Then
  \begin{equation*}
    \vect{\cM}_{ab} = \big(\cJ_{ab},\vect{\cH}_{ab}\big)^{T} \equiv \Phi^{+}(\vect{\cM}_{a})+\Phi^{-}(\vect{\cM}_{b})\in\cR.
  \end{equation*}
  \label{lem:explicitStep}
\end{lemma}
\begin{proof}
  It is straightforward to rewrite the components of $\vect{\cM}_{ab}$ as
  \begin{equation*}
    \cJ_{ab}=\f{1}{4\pi}\int_{\bbS^{2}}f_{ab}(\omegaNu)\,d\omegaNu
    \quad\text{and}\quad
    \vect{\cH}_{ab}=\f{1}{4\pi}\int_{\bbS^{2}}f_{ab}(\omegaNu)\,\vect{\ell}(\omegaNu)\,d\omegaNu,
  \end{equation*}
  where $f_{ab}(\omegaNu)=\vartheta\,f_{a}(\omegaNu)+(1-\vartheta)\,f_{b}(\omegaNu)$ and $\vartheta(\omegaNu)=(1+\widehat{\vect{e}}\cdot\vect{\ell}(\omegaNu))/2\in[0,1]$.  
  Then, since $f_{ab}(\omegaNu)\in(0,1)\,\forall\,\omegaNu\in\bbS^{2}$, we have from Lemma~\ref{lem:MomentRealizable} that $\vect{\cM}_{ab}\in\cR$.  
\end{proof}

\begin{lemma}
  Let $\vect{\cM}_{a}=(\cJ_{a},\vect{\cH}_{a})^{T}\in\cR$ and $\alpha>0$.  
  Let $\vect{\cM}_{b}=(\cJ_{b},\vect{\cH}_{b})^{T}$ satisfy
  \begin{equation*}
    \vect{\cM}_{b}=\vect{\cM}_{a}+\alpha\,\vect{\cC}(\vect{\cM}_{b}), 
  \end{equation*}
  where $\vect{\cC}(\vect{\cM})=\vect{\eta}-\vect{\cD}\,\vect{\cM}$ is the collision term in Eq.~\eqref{eq:collisionTermMoments}.  
  Then $\vect{\cM}_{b}\in\cR$.  
  \label{lem:implicitStep}
\end{lemma}
\begin{proof}
  Solving for $\vect{\cM}_{b}$ gives $\vect{\cM}_{b} = \big(\vect{I}+\alpha\,\vect{\cD}\big)^{-1}\big(\vect{\cM}_{a}+\alpha\,\vect{\eta}\big)$.  
  The first component of $\vect{\cM}_{b}$ can be written as
  \begin{equation*}
    \cJ_{b}=\f{1}{4\pi}\int_{\bbS}f_{b}(\omegaNu)\,d\omegaNu,
  \end{equation*}
  where $f_{b}(\omegaNu)=\zeta\,f_{a}(\omegaNu)+(1-\zeta)\,f_{0}$, $\zeta=1/(1+\alpha\,\xi)\in[0,1]$, and $f_{0}$ and $\xi$ are defined in Section~\ref{sec:model}.  
  Then, since $f_{b}(\omegaNu)\in(0,1)\,\forall\,\omegaNu\in\bbS^{2}$, $\cJ_{b}\in(0,1)$.  
  Similarly, we can write
  \begin{equation*}
    \vect{\cH}_{b}=\f{(1+\alpha\,\xi)}{(1+\alpha)}\,\widetilde{\vect{\cH}}_{b},
    \quad\text{where}\quad
    \widetilde{\vect{\cH}}_{b}=\f{1}{4\pi}\int_{\bbS^{2}}f_{b}(\omegaNu)\,\vect{\ell}(\omegaNu)\,d\omegaNu.  
  \end{equation*}
  It follows from Lemma~\ref{lem:MomentRealizable} that $\widetilde{\vect{\bcM}}_{b}=(\cJ_{b},\widetilde{\vect{\cH}}_{b})^{T}\in\cR$.  
  Then, since $0\le\xi\le1$, we have $|\vect{\cH}_{b}|\le|\widetilde{\vect{\cH}}_{b}| < (1-\cJ_{b})\,\cJ_{b}$.  
\end{proof}

>>>>

Our goal is to simulate massless fermions (e.g., neutrinos) and study their interactions with matter.  
The principal objective is to obtain the fermionic distribution function $f$ (or moments of $f$ as in the two-moment model employed here).  
The Pauli exclusion principle requires the distribution function to satisfy the condition $0 \le f \le 1$, which puts restrictions on the admissible values for the moments of $f$.  
In this paper, we seek to design a numerical method for solving the system of moment equations given by Eq.~\eqref{eq:momentEquations} that preserves realizability of the moments; i.e., the moments evolve within the set of admissible values as dictated by Pauli's exclusion principle.  
(Since we are only concerned with the angular dependence of $f$ in this section, we simplify the notation by suppressing the $\vect{z}$ and $t$ dependence and write $f(\omega,\vect{z},t)=f(\omega)$.)

We begin with the following definition of moment realizability.  
\begin{define}
  The moments $\vect{\cM}=\big(\cJ,\vect{\cH}\big)^{T}$ are realizable if they can be obtained from a distribution function satisfying $0 < f(\omega) < 1~\forall~\omega\in\bbS^{2}$.
  \label{def:momentRealizability}
\end{define}
\begin{rem}
  As in \cite{lareckiBanach_2011}, in Definition~\ref{def:momentRealizability}, and in the rest of this paper, we exclude the cases $f=0$ and $f=1$ almost everywhere (a.e.) on $\bbS^{2}$, which would give $\cJ=0$ and $\cJ=1$, respectively.  
\end{rem}

Realizable moments satisfy algebraic constraints.  
We proceed by stating the constraints on the moments by restating results from Theorem~7.1 in \cite{banachLarecki_2017a} in the following Lemma.  
The proof is given in \cite{banachLarecki_2017a}.  
(See also \cite{lareckiBanach_2011,banachLarecki_2013}.)
\begin{lemma}
  Suppose that $0 < f(\omega) < 1~\forall~\omega\in\bbS^{2}$.  
  Then the moments $\vect{\cM}=(\cJ,\vect{\cH})^{T}$, defined as in Eq.~\eqref{eq:angularMoments}, satisfy $0 < \cJ < 1$ and $\big(1-\cJ\big)\,\cJ-|\vect{\cH}| > 0$. 
  \label{lem:MomentRealizable} 
\end{lemma}
%\begin{proof}
%  First, let $\mu=\cos\thetaNu$ and $\bbS^{1}:=\{\,\mu\,|\,\mu\in[-1,1]\,\}$, and write
%  \begin{equation}
%    \cJ=\f{1}{2}\int_{\bbS^{1}}\mathfrak{f}(\mu)\,d\mu,
%    \quad\text{where}\quad
%    \mathfrak{f}(\mu)=\f{1}{2\pi}\int_{0}^{2\pi}f(\mu,\phiNu)\,d\phiNu.  
%  \end{equation}  
%  Since $f(\omega)\in[0,1]$, we have $\mathfrak{f}(\mu)\in[0,1]$, which implies $\cJ\in[0,1]$.  
%  Next we need to show that $|\vect{\cH}|\le(1-\cJ)\,\cJ$.  
%  Let $\widehat{\vect{n}}\in\bbR^{3}$ (independent of $\omega$) be an arbitrary unit vector, and write
%  \begin{equation}
%    \widehat{\vect{n}}\cdot\vect{\cH}
%    =\f{1}{2} \int_{\bbS^{1}}\mathfrak{f}(\mu)\,\mu\,d\mu,
%    \label{eq:NdotH}
%  \end{equation}
%  where we have aligned the polar axis of the spherical momentum space coordinate system with $\widehat{\vect{n}}$; i.e., $\widehat{\vect{n}}\cdot\vect{\ell}(\omega)=\mu$.  
%  First, following \cite{banachLarecki_2017a}, we can write the integral on the right-hand side of Eq.~\eqref{eq:NdotH} as
%  \begin{equation*}
%    (1-\cJ)\cJ - \f{1}{2}\int_{-1}^{1-2\cJ}(1-2\cJ-\mu)\,\mathfrak{f}(\mu)\,d\mu
%    -\f{1}{2}\int_{1-2\cJ}^{1}(2\cJ-1+\mu)\,[1-\mathfrak{f}(\mu)]\,d\mu.  
%  \end{equation*}
%  Since $\mathfrak{f}\in[0,1]$, $(1-2\cJ-\mu)$ is nonnegative on the interval $\mu\in[-1,1-2\cJ]$, and $(2\cJ-1+\mu)$ is nonnegative on the interval $\mu\in[1-2\cJ,1]$, we have $\widehat{\vect{n}}\cdot\vect{\cH}\le(1-\cJ)\,\cJ$.  
%  In a similar manner, we can also write the integral on the right-hand side of Eq.~\eqref{eq:NdotH} as
%  \begin{equation*}
%    -(1-\cJ)\cJ-\f{1}{2}\int_{-1}^{2\cJ-1}(1-2\cJ+\mu)\,[1-\mathfrak{f}(\mu)]\,d\mu
%    -\f{1}{2}\int_{2\cJ-1}^{1}(2\cJ-1-\mu)\,\mathfrak{f}(\mu)\,d\mu.  
%  \end{equation*}
%  Then, since $\mathfrak{f}\in[0,1]$, $(1-2\cJ+\mu)$ is nonpositive on the interval $\mu\in[-1,2\cJ-1]$, and $(2\cJ-1-\mu)$ is nonpositive on the interval $\mu\in[2\cJ-1,1]$, we have $\widehat{\vect{n}}\cdot\vect{\cH}\ge-(1-\cJ)\,\cJ$.  
%  We therefore have $-(1-\cJ)\,\cJ\le\widehat{\vect{n}}\cdot\vect{\cH}\le(1-\cJ)\,\cJ$.  
%  Since this holds for any unit vector $\widehat{\vect{n}}\in\bbR^{3}$, we must have $|\vect{\cH}|\le(1-\cJ)\,\cJ$.  
%\end{proof}

@@

For the realizability-preserving scheme developed in Section~\ref{sec:realizableDGIMEX}, we state some additional results.  
Lemma~\ref{lem:explicitStep} is used to help prove the realizability-preserving property of explicit steps in the IMEX scheme, while Lemmas~\ref{lem:implicitStep} and \ref{lem:correctionStep} are used to prove realizability-preserving properties of implicit steps.  
\begin{lemma}
  Let $\big\{\cJ_{a},\vect{\cH}_{a},\vect{\cK}_{a}\big\}$ and $\big\{\cJ_{b},\vect{\cH}_{b},\vect{\cK}_{b}\big\}$ be moments defined as in Eq.~\eqref{eq:angularMoments} with distribution functions $f_{a}$ and $f_{b}$, respectively, such that $f_{a}(\omega),f_{b}(\omega)\in(0,1)\,\forall\,\omega\in\bbS^{2}$.  
  Let $\Phi^{\pm}(\vect{\cM})=\f{1}{2}\big(\vect{\cM}\pm\widehat{\vect{e}}\cdot\vect{\cF}(\vect{\cM})\big)$, where $\widehat{\vect{e}}\in\bbR^{3}$ is an arbitrary unit vector, and $\widehat{\vect{e}}\cdot\vect{\cF}(\vect{\cM})=\big(\widehat{\vect{e}}\cdot\vect{\cH},\widehat{\vect{e}}\cdot\vect{\cK}\big)^{T}$.  
  Then
  \begin{equation*}
    \vect{\cM}_{ab} = \big(\cJ_{ab},\vect{\cH}_{ab}\big)^{T} \equiv \Phi^{+}(\vect{\cM}_{a})+\Phi^{-}(\vect{\cM}_{b})\in\cR.
  \end{equation*}
  \label{lem:explicitStep}
\end{lemma}
\begin{proof}
  It is straightforward to rewrite the components of $\vect{\cM}_{ab}$ as
  \begin{equation*}
    \cJ_{ab}=\f{1}{4\pi}\int_{\bbS^{2}}f_{ab}(\omega)\,d\omega
    \quad\text{and}\quad
    \vect{\cH}_{ab}=\f{1}{4\pi}\int_{\bbS^{2}}f_{ab}(\omega)\,\vect{\ell}(\omega)\,d\omega,
  \end{equation*}
  where $f_{ab}(\omega)=\vartheta\,f_{a}(\omega)+(1-\vartheta)\,f_{b}(\omega)$ and $\vartheta(\omega)=(1+\widehat{\vect{e}}\cdot\vect{\ell}(\omega))/2\in[0,1]$.  
  Then, since $f_{ab}(\omega)\in(0,1)\,\forall\,\omega\in\bbS^{2}$, we have from Lemma~\ref{lem:MomentRealizable} that $\vect{\cM}_{ab}\in\cR$.  
\end{proof}

\begin{lemma}
  Let $\vect{\cM}_{a}=(\cJ_{a},\vect{\cH}_{a})^{T}\in\cR$ and $\alpha>0$.  
  Let $\vect{\cM}_{b}=(\cJ_{b},\vect{\cH}_{b})^{T}$ satisfy
  \begin{equation*}
    \vect{\cM}_{b}=\vect{\cM}_{a}+\alpha\,\vect{\cC}(\vect{\cM}_{b}), 
  \end{equation*}
  where $\vect{\cC}(\vect{\cM})=\vect{\eta}-\vect{\cD}\,\vect{\cM}$ is the collision term in Eq.~\eqref{eq:collisionTermMoments}.  
  Then $\vect{\cM}_{b}\in\cR$.  
  \label{lem:implicitStep}
\end{lemma}
\begin{proof}
  Solving for $\vect{\cM}_{b}$ gives $\vect{\cM}_{b} = \big(\vect{I}+\alpha\,\vect{\cD}\big)^{-1}\big(\vect{\cM}_{a}+\alpha\,\vect{\eta}\big)$.  
  The first component of $\vect{\cM}_{b}$ can be written as
  \begin{equation*}
    \cJ_{b}=\f{1}{4\pi}\int_{\bbS}f_{b}(\omega)\,d\omega,
  \end{equation*}
  where $f_{b}(\omega)=\zeta\,f_{a}(\omega)+(1-\zeta)\,f_{0}$, $\zeta=1/(1+\alpha\,\xi)\in[0,1]$, and $f_{0}$ and $\xi$ are defined in Section~\ref{sec:model}.  
  Then, since $f_{b}(\omega)\in(0,1)\,\forall\,\omega\in\bbS^{2}$, $\cJ_{b}\in(0,1)$.  
  Similarly, we can write
  \begin{equation*}
    \vect{\cH}_{b}=\f{(1+\alpha\,\xi)}{(1+\alpha)}\,\widetilde{\vect{\cH}}_{b},
    \quad\text{where}\quad
    \widetilde{\vect{\cH}}_{b}=\f{1}{4\pi}\int_{\bbS^{2}}f_{b}(\omega)\,\vect{\ell}(\omega)\,d\omega.  
  \end{equation*}
  It follows from Lemma~\ref{lem:MomentRealizable} that $\widetilde{\vect{\bcM}}_{b}=(\cJ_{b},\widetilde{\vect{\cH}}_{b})^{T}\in\cR$.  
  Then, since $0\le\xi\le1$, we have $|\vect{\cH}_{b}|\le|\widetilde{\vect{\cH}}_{b}| < (1-\cJ_{b})\,\cJ_{b}$.  
\end{proof}
